---
title: "Circadian Mood"
output:
  word_document: default
  html_document: default
date: '2022-11-21'
---

### Install missing packages
```{r packages}
install.packages('pacman') 
pacman::p_load(lmerTest, tidyverse, lme4, wesanderson, MuMIn)
```

#Get and clean BodyFeelingExp data
```{r}
#Get FaceStroopData
datadir<-"/work/285178/BodyFeelingExp_data"

#Find files
files<-list.files(datadir,pattern='^BodyFeelingExp_.+?csv',full.names=TRUE)

#Prepare an empty data frame for the data (also removes old version)
dataBFE<-data.frame()

#How many datasets were there
n_datasets_raw<-length(files)
#Prepare a variable to monitor how many datasets we keep
n_datasets<-0
#Prepare a variable to monitor how many points we originally had
n_datapoints_raw<-0

#Loop to go through all files in the list
for(iii in 1:n_datasets_raw){
  
  #remove old loaded file to not risk importing it multiple times
  if(exists('data_temp')) rm(data_temp)
  
  #Load data
  data_temp<-read.csv(files[iii])
  if(dim(data_temp)[2]==31){
    data_temp[1,6]<-data_temp[dim(data_temp)[1],6]
    data_temp<-data_temp[1,c(6,8:27)]
        if(length(colnames(dataBFE))==0){
          dataBFE=data_temp
          rm(data_temp)
          #counter to monitor included datasets
          n_datasets<-n_datasets+1
        }
        #Bind loaded data with actual data
        else {dataBFE<-rbind(dataBFE,data_temp)
          rm(data_temp)
          #counter to monitor included datasets
          n_datasets<-n_datasets+1
        }
    }
}

#A variable to monitor how many points we keep
n_datapoints<-length(dataBFE[,1])

```

### Additional preprocessing
```{r}
#Make a variable which has hour and minutes of the day as decimal variable
dataBFE$hour2<-dataBFE$hour+(dataBFE$minute)/60
```

## Sigrid's preprocessing

```{r}
# getting info about dataset
# number of participants
length(unique(dataBFE$id)) # 35 unique entries

```

```{r}
# for loop to figure out how many times each participant participated (suspecting something is off with id's)
df_final <- data.frame()

for (i in unique(dataBFE$id)) {
  name = dataBFE %>%filter(id == i) %>% select(id) %>% slice(1)
  count = filter(dataBFE, id == i) %>% count()
  
  df_temp = data_frame(name, count)
  df_final <- rbind(df_final, df_temp)
}

# some of the ids are definitely made by mistake 
```

```{r}
# fixing participants who most likely used several IDs + lower case upper case and space issues

dataBFE <- dataBFE %>% 
  mutate(
    id = tolower(id),
    id = gsub(" ", "", id),
    id = gsub("dig05", "dlg05", id),
    id = gsub("hej123", "hej12", id),
    id = gsub("www111", "www11", id),
    id = gsub("www123", "www11", id),
    id = gsub("www12", "www11", id)
  )

# NB! rerun for loop in above chunk to make an updated boxplot below 
```


```{r}
# number of particants
length(unique(dataBFE$id)) # 35 unique entries before
# now we have 25 participants

# mean number of reports 
mean(df_final$n) # mean 11.31. # mean now; 15.84
sd(df_final$n) # sd is 14.71
min(df_final$n) # min 1       # min is still 1 ofc 
max(df_final$n) # max is 56   # max now 57

# boxplot 
df_final %>% mutate(
  class = "participant") %>% 
    ggplot() +
    aes(class, n) + 
    geom_boxplot(fill = wes_palette("Royal2")[5]) + 
  labs(
    title = "Boxplot of number of reports per participant",
    x = "", 
    y = "Number of reports") + 
   theme(
    panel.background= element_rect(fill = wes_palette("Royal1")[3]), 
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
    ) 

```


## Some plotting

```{r}
library(ggplot2)
ggplot(dataBFE,aes(x=hour2,y=HeartRate))+geom_smooth()+geom_point()
ggplot(dataBFE,aes(x=hour2,y=mood))+geom_smooth()+geom_point()
ggplot(dataBFE,aes(x=hour2,y=fresh))+geom_smooth()+geom_point()
ggplot(dataBFE,aes(x=hour2,y=hunger))+geom_smooth()+geom_point()

```
# High-school trigonometry for circidian analysis

## It all starts with a circle and a triangle
```{r}
#Radius
r<-1
# Get x-values from minus 1 to plus 1
x<-seq(-1,1,0.00001)
# A circle with c(0,0) centre can be written with these two equations (following Pythagoras)
y1<-sqrt(r^2-x^2)
y2<--sqrt(r^2-x^2)
y<-c(y1,y2)
x<-c(x,x)
#Plotting the circle with sine and cosine values
pp=pi/4
plot(x,y,type='l')
lines(x=c(0,0),y=c(-1,1))
lines(x=c(0,cos(pp)),y=c(0,sin(pp)),col='darkgreen')
lines(x=c(cos(pp),cos(pp)),y=c(0,sin(pp)),col='darkblue')
text(x=c(-0.15+cos(pp)),y=c(0.5*sin(pp)),labels='sin(x)',col='darkblue')
lines(x=c(0,cos(pp)),y=c(sin(pp),sin(pp)),col='darkred')
text(x=c(0.5*cos(pp)),y=c(+0.15+sin(pp)),labels='cos(x)',col='darkred')
```

### Going beyond the circle, the sine and cosine functions can describe cycles in time

$$
y(t)=\beta{sin}(2\pi f t )
$$
$$
y(t)=\beta{cos}(2\pi f t )
$$
Beta is the amplitude (the height of the wave)
f is the frequency (the number of cycles per time unit)
t is time point



#### Plotting circadian sine and cosine waves
```{r}
#cycle frequency (in this case per hour) - one cycle per 24 hours
cf=1/24
#sample frequency per hour
fs=100
#Duration in hours
dur=24
#A time vector divided by fs
t = seq(0, dur, 1/fs)
#Make a sine wave (with amplitude =1) for each time point in t
u = sin(2*pi*cf*t)
#Make a cosine wave (with amplitude =1) for each time point in t
u2= cos(2*pi*cf*t)
#Plot the waves
plot(x=t,y=u, type='l',col='darkblue',xlab='time(hours)',ylab='Amplitude')
text(x=1+t[1000],y=-0.2+u[500],labels='y=sin(2*pi*cf*t)',col='darkblue')
lines(x=t,y=u2, type='l',col='darkred')
text(x=-1+t[500],y=-0.2+u2[500],labels='Y=cos(2*pi*cf*t)',col='darkred')
```
As can be seen, the sine function is a time shifted version of the cosine function and vice versa.
The time shift is also called phase. We can add a constant for the phase (phi) to the sine/cosine wave function.

$$
y(t)=\beta{sin}(2\pi f t + \phi)
$$

#### Plotting circadian sine wave with phase shift
```{r}
#cycle frequency (in this case per hour) - one cycle per 24 hours
cf=1/24
#sample frequency per hour
fs=100
#Duration in hours
dur=24
#A time vector divided by fs
t = seq(0, dur, 1/fs)
#a  phase shift of pi/2 radians (half a cycle) This could be any number
phi=pi/2
#Make a sine wave (with amplitude =1)
u = sin(2*pi*cf*t)
#Make a sine wave (with amplitude =1), and phase shift
u2= cos(2*pi*cf*t+phi)
#Plot the waves
plot(x=t,y=u, type='l',col='darkblue',xlab='time(hours)',ylab='Amplitude')
text(x=1+t[1000],y=-0.2+u[500],labels='y=sin(2*pi*cf*t)',col='darkblue')
lines(x=t,y=u2, type='l',lty='dashed',col='darkblue')
text(x=-1+t[500],y=0.2+u2[500],labels='Y=cos(2*pi*cf*t+phi)',col='darkblue')
```
Using the trigonometric identity

$$
{sin(A+B)}={sin(A)cos(B)}-{cos(A)sin(B)}
$$
we can rewrite the sine function (including phase) as

$$
y(t)={\beta_1{sin}(2\pi f t)}+{\beta_2{cos}(2\pi f t)}
$$
where 
$$
{\beta_1}=\beta{cos(\phi)}, {\beta_2}=-\beta{sin(\phi)}
$$
We can use the rewritten sine function in a linear regression analysis, where we estimate the best fitting B1 and B2. This will yield a composite estimate of the amplitude and the phase of the data.

This will allow us to use sine and cosine waves to model a circadian rhythm, even if we don't know when it peaks. The amplitude will be given by

$$
{\beta}=\sqrt{\beta_1^2+\beta_2^2}
$$
The phase (phi) will be given by
$$
{\phi}=atan2(\beta_1,\beta_2)
$$

#### Use the sine and cosine waves to make 24 hour oscillation predictors for the BodyFeelingExp data

```{r}
#cycle frequency (in this case per hour) - one cycle per 24 hours
cf=1/24

#Make sine and cosine waves for each time point present in the data

dataBFE$sinCirc<-sin(2*pi*cf*dataBFE$hour2)
dataBFE$cosCirc<-cos(2*pi*cf*dataBFE$hour2)

# Plot the predictors for each data point in the data
ggplot(dataBFE, aes(x=hour2,y=sinCirc))+
      geom_point(col='darkblue')+
      geom_point(aes(y=cosCirc),col='darkred')+
      ylab('sinCirc/cosCirc')

```

## Fitting 24 hour Oscillation model

#### Freshness
```{r}
library(lmerTest)

# Freshness: Simple oscillation model
modelBFEfreshCirc<-lmer(fresh~sinCirc+cosCirc+(1|id),data=dataBFE)
m_temp<-summary(modelBFEfreshCirc)
m_temp
dataBFE$freshCircFit<-m_temp$coefficients[1,1]+m_temp$coefficients[2,1]*dataBFE$sinCirc+m_temp$coefficients[3,1]*dataBFE$cosCirc

plot(x=dataBFE$hour2,y=dataBFE$freshCircFit,type='p',col=wes_palette("Royal2")[5])

# testing colours for plotting
#wes_palette("Royal2")
#wes_palette("Royal1") # [2] er rød
#wes_palette("Royal2") # [5] er grøn

# plot 24 hour rhythm
dataBFE %>% 
    ggplot() +
    aes(hour2, fresh) +
    geom_smooth(
      aes(col = "data (smoothed)"), alpha = 0.25) +
  geom_line(aes(
    hour2, freshCircFit, col = "model fit"), size = 1) +
    labs(
      title = "Figure 1: Fresh/tired score during the day (24 hour circadian)",
      x='Time of day (hours)', 
      y='Tired/Fresh score') + 
    scale_x_continuous(breaks=c(4,8,12,16,20,24)) + # changing the x-axis values 
    scale_colour_manual("",
                      values = c("data (smoothed)" = wes_palette("Royal1")[2],
                                 "model fit" = wes_palette("Royal2")[5]
                                 )) +
  theme(
    panel.background= element_rect(fill = wes_palette("Royal1")[3]), 
      legend.title = element_blank(), 
      legend.position = c(.80, .25)
    ) 
```

#### Mood
```{r}
library(lmerTest)
# Mood: Simple oscillation model
modelBFEmoodCirc<-lmer(mood~sinCirc+cosCirc+(1|id),data=dataBFE)
m_temp<-summary(modelBFEmoodCirc)
m_temp
dataBFE$CircFit<-m_temp$coefficients[2,1]*dataBFE$sinCirc+m_temp$coefficients[3,1]*dataBFE$cosCirc

plot(x=dataBFE$hour2,y=dataBFE$CircFit,type='p',col='darkgreen')

ggplot(dataBFE,aes(x=hour2,y=mood))+geom_smooth()+geom_line(aes(x=hour2,y=CircFit+mean(mood)),col='darkgreen')+labs(x='Time of day (hours)', y='Negative/Positive Mood')
```

#### HeartRate
```{r}
# Freshness: Simple oscillation model
modelBFEHeartRateCirc<-lmer(HeartRate~sinCirc+cosCirc+(1|id),data=dataBFE)
m_temp<-summary(modelBFEHeartRateCirc)
m_temp
dataBFE$HeartRateCircFit<-m_temp$coefficients[1,1]+m_temp$coefficients[2,1]*dataBFE$sinCirc+m_temp$coefficients[3,1]*dataBFE$cosCirc

plot(x=dataBFE$hour2,y=dataBFE$HeartRateCircFit,type='p',col='darkgreen')

ggplot(dataBFE,aes(x=hour2,y=HeartRate))+geom_smooth()+geom_line(aes(x=hour2,y=HeartRateCircFit),col='darkgreen')+labs(x='Time of day (hours)', y='Tired/Fresh')
```


#### Make a 12 hour oscillation model to compare with the 24 hour model

```{r}
#cycle frequency (in this case per hour) - one cycle per 24 hours
cf=1/12

#Make sine and cosine waves for each time point present in the data

dataBFE$sinCirc12<-sin(2*pi*cf*dataBFE$hour2)
dataBFE$cosCirc12<-cos(2*pi*cf*dataBFE$hour2)

# Plot the predictors for each data point in the data
ggplot(dataBFE, aes(x=hour2,y=sinCirc12))+
      geom_point(col='darkblue')+
      geom_point(aes(y=cosCirc12),col='darkred')+
      ylab('sinCirc/cosCirc')

```

## Fitting 12 hour Oscillation models

#### Freshness
```{r}
# Freshness: Simple oscillation model
modelBFEfreshCirc12<-lmer(fresh~sinCirc12+cosCirc12+(1|id),data=dataBFE)
m_temp<-summary(modelBFEfreshCirc12)
m_temp
dataBFE$freshCircFit12<-m_temp$coefficients[1,1]+m_temp$coefficients[2,1]*dataBFE$sinCirc12+m_temp$coefficients[3,1]*dataBFE$cosCirc12

plot(x=dataBFE$hour2,y=dataBFE$freshCircFit12,type='p',col='darkgreen')

# plot 12 hour rhythm
dataBFE %>% 
    ggplot() +
    aes(hour2, fresh) +
    geom_smooth(
      aes(col = "data (smoothed)"), alpha = 0.25) +
  geom_line(aes(
    hour2, freshCircFit12, col = "model fit"), size = 1) +
    labs(
      title = "Figure 2: Fresh/tired score during the day (12 hour circadian)",
      x='Time of day (hours)', 
      y='Tired/Fresh score') + 
    scale_x_continuous(breaks=c(4,8,12,16,20,24)) + # changing the x-axis values 
    scale_colour_manual("",
                      values = c("data (smoothed)" = wes_palette("Royal1")[2],
                                 "model fit" = wes_palette("Royal2")[5]
                                 )) +
  theme(
    panel.background= element_rect(fill = wes_palette("Royal1")[3]), 
      legend.title = element_blank(), 
      legend.position = c(.80, .25)
    ) 


```


#### Mood
```{r}
library(lmerTest)
# Mood: Simple oscillation model
modelBFEmoodCirc12<-lmer(mood~sinCirc12+cosCirc12+(1|id),data=dataBFE)
m_temp<-summary(modelBFEmoodCirc12)
m_temp
dataBFE$CircFit12<-m_temp$coefficients[1,1]+m_temp$coefficients[2,1]*dataBFE$sinCirc12+m_temp$coefficients[3,1]*dataBFE$cosCirc12

plot(x=dataBFE$hour2,y=dataBFE$CircFit12,type='p',col='darkgreen')

ggplot(dataBFE,aes(x=hour2,y=mood))+geom_smooth()+geom_line(aes(x=hour2,y=CircFit12),col='darkgreen')+labs(x='Time of day (hours)', y='Negative/Positive Mood')
```

#### HeartRate
```{r}
# Freshness: Simple oscillation model
modelBFEHeartRateCirc12<-lmer(HeartRate~sinCirc12+cosCirc12+(1|id),data=dataBFE)
m_temp<-summary(modelBFEHeartRateCirc12)
m_temp
dataBFE$HeartRateCircFit12<-m_temp$coefficients[1,1]+m_temp$coefficients[2,1]*dataBFE$sinCirc12+m_temp$coefficients[3,1]*dataBFE$cosCirc12

plot(x=dataBFE$hour2,y=dataBFE$HeartRateCircFit12,type='p',col='darkgreen')

ggplot(dataBFE,aes(x=hour2,y=HeartRate))+geom_smooth()+geom_line(aes(x=hour2,y=HeartRateCircFit12),col='darkgreen')+labs(x='Time of day (hours)', y='Tired/Fresh')
```
## Combining 12 and 24 hours

#### Freshness
```{r}
# Freshness: Simple oscillation model
modelBFEfreshCircBoth<-lmer(fresh~sinCirc+cosCirc+sinCirc12+cosCirc12+(1|id),data=dataBFE)
m_temp<-summary(modelBFEfreshCircBoth)
m_temp
dataBFE$freshCircFitBoth<-m_temp$coefficients[1,1]+m_temp$coefficients[2,1]*dataBFE$sinCirc+m_temp$coefficients[3,1]*dataBFE$cosCirc+m_temp$coefficients[4,1]*dataBFE$sinCirc12+m_temp$coefficients[5,1]*dataBFE$cosCirc12

plot(x=dataBFE$hour2,y=dataBFE$freshCircFitBoth,type='p',col='darkgreen')

ggplot(dataBFE,aes(x=hour2,y=fresh))+geom_smooth()+geom_line(aes(x=hour2,y=freshCircFitBoth),col='darkgreen')+labs(x='Time of day (hours)', y='Tired/Fresh')


# plot both 24h and 12h oscillation
dataBFE %>% 
    ggplot() +
    aes(hour2, fresh) +
    geom_smooth(
      aes(col = "data (smoothed)"), alpha = 0.25) +
  geom_line(aes(
    hour2, freshCircFitBoth, col = "model fit"), size = 1) +
    labs(
      title = "Figure 3: Fresh/tired score during the day (12 and 24 hour circadian)",
      x='Time of day (hours)', 
      y='Tired/Fresh score') + 
    scale_x_continuous(breaks=c(4,8,12,16,20,24)) + # changing the x-axis values 
    scale_colour_manual("",
                      values = c("data (smoothed)" = wes_palette("Royal1")[2],
                                 "model fit" = wes_palette("Royal2")[5]
                                 )) +
  theme(
    panel.background= element_rect(fill = wes_palette("Royal1")[3]), 
      legend.title = element_blank(), 
      legend.position = c(.80, .25)
    ) 

```


```{r}
# calculate R-squared
library('MuMIn')

r.squaredGLMM(modelBFEfreshCircBoth)
# marginal Rsq = 0.2429
# conditional Rsq = 0.3827
```

#### Mood
```{r}
library(lmerTest)
# Mood: Simple oscillation model
modelBFEmoodCircBoth<-lmer(mood~sinCirc+cosCirc+sinCirc12+cosCirc12+(1|id),data=dataBFE)
m_temp<-summary(modelBFEmoodCircBoth)
m_temp
dataBFE$CircFitBoth<-m_temp$coefficients[1,1]+m_temp$coefficients[2,1]*dataBFE$sinCirc+m_temp$coefficients[3,1]*dataBFE$cosCirc+m_temp$coefficients[4,1]*dataBFE$sinCirc12+m_temp$coefficients[5,1]*dataBFE$cosCirc12

plot(x=dataBFE$hour2,y=dataBFE$CircFitBoth,type='p',col='darkgreen')

ggplot(dataBFE,aes(x=hour2,y=mood))+geom_smooth()+geom_line(aes(x=hour2,y=CircFitBoth),col='darkgreen')+labs(x='Time of day (hours)', y='Negative/Positive Mood')
```
#### HeartRate
```{r}
# Freshness: Simple oscillation model
modelBFEHeartRateCircBoth<-lmer(HeartRate~+sinCirc+cosCirc+sinCirc12+cosCirc12+(1|id),data=dataBFE)
m_temp<-summary(modelBFEHeartRateCircBoth)
m_temp
dataBFE$HeartRateCircFitBoth<-m_temp$coefficients[1,1]+m_temp$coefficients[2,1]*dataBFE$sinCirc+m_temp$coefficients[3,1]*dataBFE$cosCirc++m_temp$coefficients[4,1]*dataBFE$sinCirc12+m_temp$coefficients[5,1]*dataBFE$cosCirc12

plot(x=dataBFE$hour2,y=dataBFE$HeartRateCircFitBoth,type='p',col='darkgreen')

ggplot(dataBFE,aes(x=hour2,y=HeartRate))+geom_smooth()+geom_line(aes(x=hour2,y=HeartRateCircFitBoth),col='darkgreen')+labs(x='Time of day (hours)', y='HeartRate')
```

## Model comparisons
### Is the 24 hour or the 12 hour model the best?
```{r}
#Comparing models for freshness
anova(modelBFEfreshCirc12,  modelBFEfreshCircBoth)
anova(modelBFEfreshCirc, modelBFEfreshCircBoth)

#Comparing models for mood
anova(modelBFEmoodCirc12, modelBFEmoodCircBoth)
anova(modelBFEmoodCirc, modelBFEmoodCircBoth)

#Comparing models for heartrate
anova(modelBFEHeartRateCirc12,modelBFEHeartRateCircBoth)
anova(modelBFEHeartRateCirc,modelBFEHeartRateCircBoth)

#effect(modelBFEHeartRateCirc)
```

### Activity and Mental activity
```{r}

library(ggplot2)
ggplot(dataBFE,aes(x=MentalActivity))+geom_bar()
ggplot(dataBFE,aes(x=Activity))+geom_bar()+coord_flip()

#Mood as function of Mental activity
modelBFEmoodMentalAct<-lmer(mood~MentalActivity + (1|id),data=dataBFE)
anova(modelBFEmoodMentalAct)

#Mood as function of activity
modelBFEmoodAct<-lmer(mood~Activity + (1|id),data=dataBFE)
anova(modelBFEmoodAct)
```

### Mood by month
```{r}
library(ggplot2)
ggplot(dataBFE,aes(y=mood, x=as.factor(month)))+geom_boxplot()
```


### Mood by hunger level
```{r}
model_mood_hunger <- lmer(mood ~ hunger+(1|id),data=dataBFE)
m_temp <- summary(model_mood_hunger)
m_temp

dataBFE %>% 
  ggplot() +
  aes(x=hour2, y=mood) + 
  geom_smooth() + 
  geom_smooth(aes(hour2, hunger), col = 'darkgreen') + 
  coord_cartesian(y = c(0,100))+geom_line(aes(x=hour2,y=HeartRateCircFitBoth),col='darkgreen')+labs(x='Time of day (hours)', y='HeartRate')
```

```{r}
# Demographics 
df_final <- data.frame()

for (i in unique(dataBFE$id)) {
  name = dataBFE %>%filter(id == i) %>% select(id, Gender, age, Handedness) %>% slice(1)
  count = filter(dataBFE, id == i) %>% count()
  
  df_temp = data_frame(name, count)
  df_final <- rbind(df_final, df_temp)
}

# gender
df_final %>% filter(
  Gender == "female"
  ) %>% count() # 17 girls

df_final %>% filter(
  Gender == "male"
  ) %>% count() # 8 boys

# age
mean(df_final$age) # mean age 24.44
sd(df_final$age) # sd age 5.38
min(df_final$age) # 21
max(df_final$age) # 49 

```

```{r}
# figuring out the start and end data of the experiment by manipulating the Time column
dataBFE$Time

dataBFE <- dataBFE %>% 
  mutate(
    new_date = as.Date(Time, format = "%a %b %d %Y")
  )
# start data Sep 8
# end data Nov 21 
# = 2 months and 13 days 
```

